### 2010-03-17 papers/icfp-2010-desc/paper_idesc.tex
IDesc@ICFP: done with Hutton's razor

@@

\subsection{Examples}
\label{sec:idesc-examples}

@@

@@

\begin{structure}
<- Hutton's razor
    ???
\end{structure}

\note{Begging for a transtion here.}

@@

\begin{structure}
<- Hutton's razor is a free monad
    <- What is it?
        -> It's an evaluator
    ???
\end{structure}

@@

@@

\[\stk{
\IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
\IDescD\: I \mapsto \FreeMonad{[\IDescFreeD]}{I}\: I
}\]

>>>>

\begin{wstructure}
<- Tagged constructor choice
    <- Index available when defining the data-type
        -> Can influence the choice of constructors
        -> Dependently-typed data-types
            <- term (index) influence types
    -> Two parts
        <- Always on the menu
            <- E : EnumU 
            <- ED : E -> IDesc I
        <- Index-dependent
            <- F : I -> EnumU 
            <- FD : (i : I) -> spi (F i) (\_ -> IDesc I)
        -> taggedIDesc I == Sigma E ED x Sigma F FD
        -> toIDesc : (I : Set) -> taggedIDesc I -> (I -> IDesc I)
\end{wstructure}

In Section~\ref{sec:desc-examples}, we have defined a \emph{tagged}
form for descriptions. This format follows the usual presentation of
inductive types as sums of product. With indexed description, this
definition can be generalized. Indeed, when defining an indexed
data-type, we have access to this index. Therefore, we can use this
index to influence the choice of constructors. This captures the
essence of dependent data-types: a term -- the index -- has the
ability to influence the data-type.

For convenience, we divide a tagged indexed description in two parts:
first, the constructors that do not depend on the index; then, the
constructors that do. The non-dependent part mirrors the definition
for non-indexed descriptions: we are provided a finite choice of
constructors. The index-depend part simply indexes the choice of
constructors by $I$. Hence, by inspecting the index, it is possible to
enable or disable the constructors.

\[\stk{
 \ATagIDesc{I} \mapsto \SIGMA{\V{E}}{\EnumU} (\PI{\V{i}}{I} \spi{\V{E}}{(\LAM{\_} \IDesc{I})}) \\
 \ITagIDesc{I} \mapsto \SIGMA{\V{F}}{I \To \EnumU} (\PI{\V{i}}{I} \spi{(F\: i)}{(\LAM{\_} \IDesc{I})})  \\
 \TagIDesc{I}  \mapsto \TIMES{\ATagIDesc{I}}{\ITagIDesc{I}}
}\]

In the case of vectors, for instance, for the index $n = \NatZero$, we
would only offer the constructor $\ListNil$. Similarly, for $n =
\NatSuc{n'}$, we only offer the constructor $\ListCons{}{}$. We will
see more instances of this pattern in the following.

@@

\subsubsection{Examples}
\label{sec:idesc-examples}

\paragraph{Natural numbers:}

@@

\paragraph{Indexed descriptions:}

@@

\paragraph{Typed expressions:}

\begin{wstructure}
<- Hutton's razor
    <- Types
        <- 'Nat
        <- 'Bool
    <- Term [figure]
        <- val : Val 'a -> 'a  for Val : Ty -> Set, mapping to Nat and Bool
        <- cond : 'Bool -> a -> a -> a
        <- plus : 'Nat -> 'Nat -> 'Nat
        <- le : 'Nat -> 'Nat -> 'Bool
\end{wstructure}

%% Types
\newcommand{\Ty}{\C{Ty}}
\newcommand{\Ebool}{\etag{\CN{bool}}}
\newcommand{\Enat}{\etag{\CN{nat}}}
\newcommand{\Epolym}{\etag{\alpha}}

%% Constructors
\newcommand{\Eval}[1]{\etag{\CN{val}}~#1}
\newcommand{\Econd}[3]{\etag{\CN{cond}}~#1~#2~#3}
\newcommand{\Eplus}[2]{\etag{\CN{plus}}~#1~#2}
\newcommand{\Ele}[2]{\etag{\CN{le}}~#1~#2}

%% Index mapper (terminology?)
\newcommand{\Val}[1]{\D{Val}~#1}
\newcommand{\Var}[2]{\D{Var}_{#1}~#2}

%% Hutton expressions
\newcommand{\HExprD}{\C{ExprD}}
\newcommand{\HExprAD}{\C{ExprAlwaysD}}
\newcommand{\HExprID}{\C{ExprIndexedD}}
\newcommand{\HExprVarD}{\C{ExprD}_{\C{Var}}}
\newcommand{\HExprFreeD}{\C{ExprD}_{\C{Free}}}
\newcommand{\HExprAFreeD}{\C{ExprAlwaysD}_{\C{Free}}}

So far, the examples we have seen lives in $\IDesc{\Unit}$, hence are
not using any indexing. We are going to define a syntax for a small
typed language. We consider two types, natural numbers and booleans:

\[
\Ty \mapsto \EnumT{[\Enat \quad \Ebool]}
\]

An expression of this language is either a value, a conditional
expression, addition of numbers, or comparison of numbers. Informally,
their type is the following:

\[
\begin{array}{l@{\::\:\:}l}
\Econd{}{}{}     & \Ebool \To \Epolym \To \Epolym \To \Epolym  \\ 
\Eplus{}{}       & \Enat \To \Enat \To \Enat                   \\
\Ele{}{}         & \Enat \To \Enat \To \Ebool                  \\
\Eval{}          & \Val{\Epolym} \To \Epolym
\end{array}
\]

The notation $\Epolym$ represents a form of ``polymorphism'', meaning
that $\alpha$ can be any type. \note{Polymorph? Will any ghost try to
  haunt me with that?  A ghost under a mixed prefix?
  Aaaaaaaaaaargh\ldots}

The function $\Val{}$, used in the definition of $\Eval{}$, simply
maps $\Epolym$ to the corresponding type in the host language. Hence,
the argument of $\Eval{}$ are ensured to be of the expected type. We
assume $\Nat$ and $\Bool$ represent natural numbers and booleans in
the host language. Further, we also assume the existence of an
addition and comparison operator in $\Nat$, respectively name
$\F{plusHost}$ and $\F{leHost}$. We define $\Val{}$ as follow:

\[\stk{
\Val{} : \Ty \To \Set \\
\begin{array}{@{}l@{\:=\:\:}l}
\Val{\Enat}   & \Nat \\
\Val{\Ebool}  & \Bool
\end{array}
}\]

In our universe of descriptions, this data-type is represented by a
tagged indexed description. We use the index to carry the type: the
resulting description is indexed by $\Ty$. We observe that some
constructors are ``polymorph'', namely $\Econd{}{}{}$ and $\Eval{}$. On the
other hand, the $\Eplus$ and $\Ele$ constructors are
index-dependent. $\Eplus$ is defined if and only if the result type --
the index -- is $\Enat$, whereas $\Ele$ is defined if and only if the
index is $\Ebool$. The actual code precisely follows this intuition,
as shown in Figure~\ref{fig:hexpr-full}. For brevity, we use an
informal $\case{\ldots}{\ldots}$ notation, simulating a definition by
pattern-matching. Formally, this corresponds to a call to the
$\switch{}{}{}{}$ eliminator.

\[\stkc{
\stk{
\HExprD : \TagIDesc{\Ty} \\
\HExprD \mapsto ( \HExprAD , \HExprID ) \\
} \\
\\
\begin{array}{ll}
\stk{
\HExprAD : \ATagIDesc{\Ty} \\
\HExprAD \mapsto \left[
                 \begin{array}{l}
                   \EnumT{[\Eval{}  \Econd{}{}{}]} \red{,} \\
                   \LAM{ty}
                   \left[ %%! Should be read
                   \begin{array}{l}
                   \DConst{(\Val{ty})} \\
                   \DProd{\DVar{\Ebool}}{\DProd{\DVar{ty}}{\DVar{ty}}}
                   \end{array}
                   \right]
                 \end{array}
                 \right] %%! Should be \red
\\
\\
} &
\stk{
\HExprID : \ITagIDesc{\Ty} \\
\HExprID \mapsto \left[
                 \begin{array}{l}
                   \LAM{ty} \case{ty}{\Enat \To \EnumT{[\Eplus{}{}]} \\ \Ebool \To \EnumT{[\Ele{}{}]}} \red{,} \\
                   \LAM{ty}
                   \case{ty}{
                     \Enat \To \DProd{\DVar{nat}}{\DVar{nat}} \\
                     \Ebool \To \DProd{\DVar{nat}}{\DVar{nat}}
                   }
                   \end{array}
                   \right]
}
\end{array}
}\]

\caption{Syntax of typed expressions}
\label{fig:hexpr-full}

\note{the big square brackets should be red}

\begin{wstructure}
    -> evaluation: IMu TermD -> Val
        -> it is a catamorphism
            <- Look closer at the type
        -> implementation [code]
            <- Just define one reduction step
            -> cata does the rest
                /> cata is for free!
\end{wstructure}

\newcommand{\evalH}{\F{eval}_{\green{\Downarrow}}}
\newcommand{\evalOne}{\F{eval}_{\green{\downarrow}}}

Having implemented the syntax, we would like to describe its
semantics. To do so, we implement an evaluator. The type of the
evaluator is:

\[
\evalH : \PI{\V{ty}}{\Ty} 
         \IMu{\Ty}{\toIDesc{\HExprD}}{ty} \To
         \Val{ty}
\]

The type of $\F{eval}$ is strikingly similar to a
catamorphism. Indeed, implementing a single step of evaluation -- the
algebra -- is sufficient, as $\F{cataI}$ gives, for free, the
full evaluator. The implementation is as follow:

\[\stk{
\evalOne : \PI{\V{ty}}{\Ty} \idescop{\toIDesc{\HExprD}}{\Ty}{ty} \To {\Val{ty}} \\
\begin{array}{@{}l@{\:=\:\:}l}
\evalOne\: \_\: \pair{\Eval{}}{x}{}                                         & x \\
\evalOne\: \_\: \pair{\Econd{}{}{}}{\pair{\BoolTrue}{\pair{x}{\_}{}}{}}{}   & x \\
\evalOne\: \_\: \pair{\Econd{}{}{}}{\pair{\BoolFalse}{\pair{\_}{y}{}}{}}{}  & x \\
\evalOne\: \Enat\: \pair{\Eplus{}{}}{\pair{x}{y}{}}{}                       & \F{plusHost}\: x y \\
\evalOne\: \Ebool\: \pair{\Ele{}{}}{\pair{x}{y}{}}{}                        & \F{leHost}\: x y 
\end{array} \\
\\
\evalH : \PI{\V{ty}}{\Ty} 
           \IMu{\Ty}{\toIDesc{\HExprD}}{ty} \To
           \Val{ty} \\
\evalH\: ty\: term = \F{cataI}\: \Ty\: 
                                 \toIDesc{\HExprD}\: 
                                 \Val{}\: 
                                 \evalOne\: 
                                 ty\: 
                                 term
}\]

\note{We have not defined \F{cataI}!}

\begin{wstructure}
    /> Closed term
        <- only constants and operations on them
        -> Extend Val with Var : Ty -> Set, mapping to EnumU
            -> Open term
            -> Language of well-typed terms
                <- By construction
\end{wstructure}

Hence, we have defined the syntax of a typed language of arithmetic
and boolean operations. We have given its semantics through an
evaluation function. Provided a one step semantic of the language, the
big step evaluation is granted without effort thanks to the generic
catamorphism.

However, so far, we are only able to define and manipulate
\emph{closed} terms. By abstracting over $\Var{}$, it is possible to
build and manipulate \emph{open} terms, that is terms with symbolic
variables. On the model of $\Val{}$, we define $\Var{}{}$:

\[\stk{
\Var{}{} : \EnumU \To \Ty \To \Set \\
\Var{dom}{\_} = \EnumT{dom}
}\]

Whereas $\Val{}$ was mapping the type to the corresponding host type,
$\Var{}{}$ maps types to a finite set. The finite set -- the context
-- contains values. A variable is therefore a $\Eval{}$ that contains
a pointer to a particular element of the finite set -- an element of
$\EnumT{dom}$. The extra argument to $\Var{}{}$ is the domain of this
context.

Consequently, replacing $\Val{ty}$ by $(\SUM{\Val{ty}}{\Var{dom}{ty}})$ in
Figure~\ref{fig:hexpr-full} turns the language of closed term into a
language of opened terms with constants. \note{Is it polite and
  comprehensible to ask that?} This defines a new indexed description,
called $\HExprVarD$.

\begin{wstructure}
        <- evaluator, with a context
            -> First, close variables
                -> Perform assignment
                <- subst [code]
\end{wstructure}

\newcommand{\discharge}{\F{discharge}}

Again, we would like to give a semantics to this extended language. We
proceed in two steps: first, replace the variables by their value in
the context; then, evaluate the resulting closed term. Thanks to
$\evalH$, we are already able to solve the second problem. Let us
focus on discharging variables from the context. Again, we can
subdivide this problem: first, we need the ability to discharge a
single variable from the context; then, we apply this $\discharge$
function on every variables in the term.

The $\discharge$ function is relative to the required type, the domain
of the context, and a context containing values of the corresponding
type. Its action is to map variables to their value in context, and
directly return constant values. This corresponds to the following
function:

\[\stk{
\begin{array}{@{}ll}
\discharge : & \PITEL{\V{ty}}{\Ty}
               \PITEL{\V{dom}}{\EnumU} \\
             & \PI{\V{\gamma}}{\spi{dom}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD}}{ty})}} \\
             & (\SUM{\Val{ty}}{\Var{dom}{ty}}) \To
               \IMu{\Ty}{\toIDesc{\HExprD}}{ty} 
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\discharge\: ty\: dom\: \gamma\: (\SumLeft x)  & \Con{\pair{\Eval{}}{x}{}} \\
\discharge\: ty\: dom\: \gamma\: (\SumRight v) & \switch{dom}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD}}{ty})}{\gamma}{v}
\end{array}
}\]

\begin{wstructure}
            /> Then, perform subst everywhere in the term
                -> Show type [code]
                /> This is a bind!?
                -> There is some more structure 
                    -> We should try to get it
\end{wstructure}

Having implemented the local $\discharge$ operation, we are left with
applying it over all variables of the term. The type of this operation
is the following:

\newcommand{\substH}{\F{substExpr}}

\[
\begin{array}{@{}ll}
\substH  : & \PITEL{\V{ty}}{\Ty}
            \PITEL{\V{dom}}{\EnumU} 
            \PITEL{\V{\gamma}}{\spi{dom}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD}}{ty})}} \\
          &  \PI{\V{\sigma}}{\PI{\V{ty}}{\Ty} (\Val{ty} + \Var{dom}{ty}) \To
                             \IMu{\Ty}{(\toIDesc{\HExprD})}{ty}} \\
          & \IMu{\Ty}{\toIDesc{\HExprVarD}}{\V{ty}} \To
            \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}}
\end{array}
\]

Abstracting away the book-keeping introduced by the context, this
definition looks familiar. Indeed, it is extremely similar to a
monadic ``bind''. This is not surprising as we are defining a
first-order syntax with variables: our data-type enjoys more structure
than what we are currently provided. In particular, this is
reminiscent to a free monad, where $\Eval{}$ is the ``return''
introducing variables. Instead of implementing $\substH$ in this
special case, we shall implement it in a generic setting, and express
our example in that setting.

@@

\subsubsection{Examples}

\paragraph{Typed expressions:}

\begin{wstructure}
<- Hutton's razor is a free monad
    <- substI was our candidate bind
        -> Massage the definition of expr to get it for free
    <- Finding the functor
        <- 'val is the return
        <- The other components are the action
        -> Updated tagged description [figure]
    -> Compute the free monad Hutton * X
        -> With X = Val: We get back our closed terms
        -> With X = Val + Var vars: We get back our open terms
\end{wstructure}

In Section~\ref{sec:idesc-examples}, we had the intuition that our
datatype $\HExprD$ and $\HExprVarD$ enjoys a monadic structure. We had
identified the variable substitution operation as the ``bind'' of a
free monad. The definition of $\substI$ above confirm our intuition:
using $substI$, we should easily obtain $\substH$. To do so, we first
have to massage the definition of our data-type, to exhibit its
monadic structure.

As previously mentioned, we identify $\Eval{}$ as the ``return'' of
the free monad, while the other components are the action of the
monad. As a result, the definition is similar to $\HExprD$ presented
in Figure~\ref{fig:hexpr-full}, at the exception of $\HExprAD$ that is
replaced by $\HExprAFreeD$:

\[\stk{
\HExprAFreeD : \ATagIDesc{\Ty} \\
\HExprAFreeD \mapsto \left[
                 \begin{array}{l}
                   \EnumT{[\Econd{}{}{}]} \red{,} \\
                   \LAM{ty}
                   \left[ %%! Should be red
                   \begin{array}{l}
                   \DProd{\DVar{\Ebool}}{\DProd{\DVar{ty}}{\DVar{ty}}}
                   \end{array}
                   \right]
                 \end{array}
                 \right] %%! Should be \red
}\]

The resulting data-type is called $\HExprFreeD$. By a simple unfolding
of definition, we note that $\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}$
corresponds to the closed expression, $\HExprD$. Similarly,
$$\FreeIMonad{\HExprFreeD}{\Ty}{(\LAM{ty} \Val{ty} + \Var{ty})}$$
corresponds to expressions with variables, $\HExprVarD$.

\begin{wstructure}
    /> On open terms, we get a substitution
        -> Apply substI on assgnmt
            -> Have a well-typed language
            -> Get a safe evaluator 
                <- for well-typed terms 
                <- in well-typed contexts
\end{wstructure}

The evaluator for closed term we implemented in
Section~\ref{sec:idesc-examples} remains unchanged. It reduces closed
terms in $\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}$ to values in
$\Val{ty}$. We are therefore left with implementing $\substH$. We
simply have to fill in the right arguments to $\substI$, the type
guiding us:

\[\stk{
\begin{array}{@{}ll}
\substH  : & \PITEL{\V{ty}}{\Ty}
             \PITEL{\V{dom}}{\EnumU} 
             \PITEL{\V{\gamma}}{\spi{dom}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}}}{ty})}} \\
           &  \PI{\V{\sigma}}{\PI{\V{ty}}{\Ty} (\Val{ty} + \Var{dom}{ty}) \To
                              \IMu{\Ty}{(\toIDesc{\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}})}{ty}} \\
           & \IMu{\Ty}{\toIDesc{\FreeIMonad{\HExprFreeD}{\Ty}{\Val{} + \Var{dom}{}}}}{\V{ty}} \To
             \IMu{\Ty}{\toIDesc{\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}}}{\V{ty}} \\
\end{array} \\
\substH\: ty\: dom\: \gamma\: \sigma\: term \mapsto 
    \substI\: \Ty\: 
              (\Val{} + \Var{dom}{})\: 
              \Val{}\:
              \HExprFreeD\:
              \sigma\:
              ty\:
              term
}\]

\note{Not using $\discharge$, that's a shame.}

Hence, we have completed our implementation of the evaluator for open
terms. We started with a well-typed language of arithmetical
expressions. We have seen how to take advantage of indexes to only
present type-safe constructors. Then, we have implemented an evaluator
for closed term, based on the generic catamorphism function. Having
implemented an open term representation, we wanted a substitution
operation, in order to close open terms in a context. Again, we have
implemented this operation with the generic substitution
operator. Hence, without much efforts, we have described the syntax of
a well-typed language, together with its semantics.

@@

\paragraph{Indexed description:}

@@

\[\stk{
\IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
\IDescD\: I \mapsto \FreeIMonad{[\IDescFreeD]}{\Unit}{I}\: I
}\]
